Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-07 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
Section III.A

Having reviewed the modified gravity theories we will consider, we now present
the frameworks upon which we will build our waveform model to confront these
theories against GW observations.
%
We will start by reviewing the theory-agnostic parametrized ringdown
spin-expansion coefficients (ParSpec) expansion in Sec.~\ref{sec:review_parspec}.
%
Next, in Sec.~\ref{sec:review_pSEOB}, we will review the
inspiral-(plunge)-merger-ringdown (IMR) BBH waveform
model~\cite{Brito:2018rfr,Ghosh:2021mrv} we will use and explain how we modify
it to include the ParSpec expansion.
%
Finally, in Sec.~\ref{sec:theory_specific_qnm}, we show how we can map
theory-specific QNM calculations in modified gravity theories onto the
theory-agnostic free coefficients in the ParSpec expansion.
%
Ultimately, this provides us with an IMR BBH waveform model, with the ringdown
portion of the model informed by QNM calculations in specific beyond-GR
theories.

@@

We start by reviewing the ParSpec
framework introduced in Ref.~\cite{Maselli:2019mjd}, but following
some of the notation used in Ref.~\cite{Carullo:2021dui}.
%
A general procedure to describe deviations to the QNM frequencies $\omega_{k}^{\rm Kerr}$ and
damping times $\tau_{k}^{\rm Kerr}$ of the GR Kerr BH is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{k} &= \omega^{\rm Kerr}_{k} \, (1 + \delta \omega_{k}), \\
\tau_{k}   &= \tau^{\rm Kerr}_{k}   \, (1 + \delta \tau_{k}),
\end{align}
\end{subequations}
%
where $\delta\omega_{k}$ and $\delta\tau_{k}$ are the deformation parameters.
%
Here we use the subscript $k$ to represent the QNM labels $\{l,\, m,\, n\}$,
where $l$ and $m$ are the multipole indices, and $n$ the overtone number.
%
This type of parametrization was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Carullo:2018sfu}.

A caveat of parametrizations of the form~\eqref{eq:general_deviation}, is that
$\delta\omega_{k}$ and $\delta\tau_{k}$ depend, in general, on the BH's source mass and spin.
%
Ideally, one would like to explicitly reinstate the dependence the source-frame
BH mass and spin for it would
%
(i) introduce deformation parameters which can be determined, once and for
all, from a specific gravity theory (GR included), and
%
(ii) make it simpler to combine constraints coming from multiple source.

In the ParSpec expansion this is accomplished by performing a bivariate
expansion of Eq.~\eqref{eq:general_deviation} in the detector-frame BH's mass
$M$ and spin $\chi$, namely
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{k} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{k} \, \left( 1 + \gamma \delta \omega^{(n)}_{k} \right), \\
\tau_{k}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{k}   \, \left( 1 + \gamma \delta \tau^{(n)}_{k} \right),
\end{align}
\end{subequations}
%
where $\omega_{k}^{n}$ and $\tau_{k}^{n}$ are dimensionless coefficients of the
expansion in spin for Kerr QNMs;
%
$\delta\omega_{k}^{(n)}$ and $\delta\tau_{k}^{(n)}$ are source-independent
dimensionless coefficients that characterize the corrections to the Kerr QNM at
each spin-order;
%
and finally $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, given by
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on a parameter $\ell$, with dimensions of length (and sets the
lengthscale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\ell$ dimensionless by using the other
lenghtscale associated to the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$, which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}.

In principle, a modification to GR would also affect $M$ and $\chi$ and the
expansion should be written in terms of non-GR mass and spin, say $\bar{M}$ and
$\bar{\chi}$. If we assume that the beyond-GR corrections are included perturbatively
(as is the case for all theories described in Sec.~\ref{sec:review_theories}),
the modifications to the BH mass and spin can be absorbed into the deviations
parameters $\delta\omega$ and $\delta\tau$.
%
This means we can identify the $M$ and $\chi$ with their \emph{corresponding GR values}.
%
We will see that in Sec.~\ref{sec:pe} that this is assumption is indeed
satisfied in our parameter estimation studies (cf.~Fig.~\ref{fig:corner_plot}).

Finally, we remark that in the GR limit ($\gamma \to 0$)
the series~\eqref{eq:kerr_expansion} truncated at $N_{\rm max} = 4$, reproduces with 1\% accuracy
the Kerr QNMs for spins $\chi \lesssim 0.7$.
%
The values of the fitting coefficients $\omega_{k}^{(n)}$ and $\tau_{k}^{(n)}$
can be found in Ref.~\cite{Maselli:2019mjd}, Table I.
%
In Ref.~\cite{Carullo:2021dui}, the fitting coefficients were calculated up to $N_{\rm max} = 9$,
which extend the validity of the spin-expansion up to $\chi \lesssim 0.99$.
%
As we will discuss in Sec.~\ref{sec:pe}, the expansion to $N_{\rm max} = 4$ is
sufficient for our purposes.

>>>>

Having reviewed the modified gravity theories we will consider, we now present
the sequential building blocks for the waveform model we use to test these theories
against GW observations.
%
We start by reviewing the theory-agnostic parametrized ringdown
spin-expansion coefficients (ParSpec) \sout{expansion} \ag{framework} \agcomm{replaced `framework' to prevent duplication.  The original parspec authors call this a data-analysis framework, so I think it would be safe to use it. I have left the term ``ParSpec expansion" in other places, because ParSpec is always abbreviated in those places.} in Sec.~\ref{sec:review_parspec}.
%
Next, in Sec.~\ref{sec:review_pSEOB}, we review our baseline
parametrized inspiral-(plunge)-merger-ringdown (IMR) BBH waveform
model~\cite{Brito:2018rfr,Ghosh:2021mrv} and explain how we modify
it to include the ParSpec expansion.
%
Finally, in Sec.~\ref{sec:theory_specific_qnm}, we show how we can map
theory-specific QNM calculations in modified gravity theories onto the
theory-agnostic free coefficients in the ParSpec expansion.
%
Ultimately, this provides us with an IMR BBH waveform model, with the ringdown
portion of the model informed by QNM calculations in specific beyond-GR
theories.

@@

We start by reviewing the ParSpec
framework introduced in Ref.~\cite{Maselli:2019mjd}, but following
some of the notation used in Ref.~\cite{Carullo:2021dui}.
%
A general procedure to describe deviations to the QNM frequencies $\omega_{k}^{\rm Kerr}$ and
damping times $\tau_{k}^{\rm Kerr}$ of the GR Kerr BH is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{k} &= \omega^{\rm Kerr}_{k} \, (1 + \delta \omega_{k}), \\
\tau_{k}   &= \tau^{\rm Kerr}_{k}   \, (1 + \delta \tau_{k}),
\end{align}
\end{subequations}
%
where $\delta\omega_{k}$ and $\delta\tau_{k}$ are the fractional deformation parameters.
%
Here we use the subscript $k$ to represent the QNM labels $\{l,\, m,\, n\}$,
where $l$ and $m$ are the multipole indices, and $n$ the overtone number.
%
This type of parametrization was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Carullo:2018sfu}.

\ag{The current LIGO-Virgo-Kagra tests of BH ringdown (see Sec.VII.A of~\cite{LIGOScientific:2020tif} or Sec.VIII.A of~\cite{LIGOScientific:2021sio}) take a restrictive theory-agnostic approach towards the inference of $\delta\omega_{k}$ and $\delta\tau_{k}$. These deviations are either assumed to occur identically across all observed sources or belong to a generic underlying Gaussian population. A more realistic scenario, however, is that these parameters depend on the source BH's mass and spin.}
\sout{A caveat of parametrizations of the form~\eqref{eq:general_deviation}, is that
$\delta\omega_{k}$ and $\delta\tau_{k}$ depend, in general, on the BH's source mass and spin.}
%
Ideally, one would like to explicitly reinstate \sout{the dependence the source-frame
BH mass and spin} \ag{this dependence} for it would
%
(i) introduce deformation parameters which can be determined, once and for
all, from a specific gravity theory (GR included), and
%
(ii) make it simpler to combine constraints coming from multiple \ag{(independently observed)} sources.

ParSpec is such an observable-based
\sout{In the ParSpec expansion this is accomplished by performing a} bivariate
expansion of Eq.~\eqref{eq:general_deviation} in the detector-frame BH's mass
$M$ and spin $\chi$, \sout{namely}
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{k} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{k} \, \left( 1 + \gamma \delta \omega^{(n)}_{k} \right), \\
\tau_{k}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{k}   \, \left( 1 + \gamma \delta \tau^{(n)}_{k} \right),
\end{align}
\end{subequations}
%
where $\omega_{k}^{(n)}$ and $\tau_{k}^{(n)}$ are dimensionless coefficients of the
expansion in spin for Kerr QNMs;
%
$\delta\omega_{k}^{(n)}$ and $\delta\tau_{k}^{(n)}$ are source-independent
dimensionless coefficients that characterize the corrections to the Kerr QNM at
each spin-order;
%
and $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, given by
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on a parameter $\ell$, with dimensions of length (and sets the
length-scale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made \sout{$\ell$} \ag{$\gamma$} dimensionless  by using the other
length-scale associated with the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$, which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}. \ag{Also, dividing by the factor $G/c^2$ allows us to
express $\ell$ in physically intuitive metric units.}

In principle, a modification to GR would also affect $M$ and $\chi$ and the
expansion should be written in terms of the non-GR mass and spin, say $\bar{M}$ and
$\bar{\chi}$. If we assume that the beyond-GR corrections are included perturbatively
(as is the case of all theories described in Sec.~\ref{sec:review_theories}),
the modifications to the BH mass and spin can be absorbed into the deviations
parameters $\delta\omega$ and $\delta\tau$.
%
This means we can identify the $M$ and $\chi$ with their \emph{corresponding GR values}
\footnote{\ag{For a more detailed discussion, see Appendix A in ~\cite{Maselli:2019mjd}.}}.
%
We will see that in Sec.~\ref{sec:pe} that this assumption is indeed
satisfied in our parameter estimation studies (cf.~Fig.~\ref{fig:corner_plot}).

Finally, we remark that in the GR limit ($\gamma \to 0$)
the series~\eqref{eq:kerr_expansion} truncated at $N_{\rm max} = 4$, reproduces with 1\% accuracy
the Kerr QNMs for spins $\chi \lesssim 0.7$.
%
The values of the fitting coefficients $\omega_{k}^{(n)}$ and $\tau_{k}^{(n)}$
can be found in Ref.~\cite{Maselli:2019mjd}, Table I\agcomm{This sentence would need
to be adapted if we put the GR coefficients in our current Table II.}.
%
In Ref.~\cite{Carullo:2021dui}, the fitting coefficients were calculated up to $N_{\rm max} = 9$,
which extend the validity of the spin-expansion up to $\chi \lesssim 0.99$.
%
As we will discuss in Sec.~\ref{sec:pe}, the expansion to $N_{\rm max} = 4$ is
sufficient for our purposes.
